Causal Forest Approach for Site-specific Input Management via On-farm Precision Experimentation

### 2022-02-07 Writing/manuscript_cea.rmd
changes in appendix and manuscript

\begin{equation}
f(N_{i}: \alpha_{i,j}, \beta_{i,j}, ymax_{i,j}) = ymax_{i,j}(1-exp(\alpha_{i,j} + \beta_{i,j} \cdot N_{i})) + \varepsilon^{i,j}. (\#eq:eqn6)
\end{equation}

The first term on the right-hand side of Eq. \@ref(eq:eqn6) is the deterministic yield component. The variable $N$ represents the nitrogen fertilizer application rate (kg/ha) which the trial designs varied by plot. Also in Eq. \@ref(eq:eqn6), $\alpha$, $\beta$, and $ymax$ are spatially varying parameters that govern how yield level is determined in conjunction with the applied N rate. They reflect local cell-level production conditions, such as soil characteristics and topography. For example, $\alpha$ and $ymax$, which affect yield independently from the applied N fertilizer rate, might be interpreted as depending on the ingredient soil N content supplied by residual N, soil organic matter, etc. $\beta$ can be interpreted as reflecting how efficiently the crop uses N to create grain mass, and might depend on soil properties that affect the availability of applied N fertilizer to the crop, say due to N immobilization, leaching, denitrification, and volatilization [@johnson2005nitrogen; @alva2006nitrogen]. The error term of the yield $\varepsilon$ represents the composite of all the unobserved factors that affect crop yield beyond $\alpha$, $\beta$, $ymax$ and $N$. Each cell $(i,j)$'s "true" optimal N application rate, denoted $N^{opt}_{i,j}$, was calculated by solving a profit maximization problem by cell assuming that the prices of corn and N fertilizer were $\$0.138/kg$ and $\$1.323/kg$, respectively. These prices are used consistently for our simualtions. See Appendix \@ref(par-gen) for further details on how the paremeters were generated and sample maps of the parameters and true economically optimal N rates.

@@

In subsequent modeling scenarios, we made the simulated experimental data more realistic by varying the degree to which the researcher understood what the needed covariates' values were those in the data set. In the second scenario, denoted "abytt," we introduced additional variables $\theta^1$ and $\theta^2$, and the researcher's data set contained values of $\alpha_{i}$, $\beta_{i}$, $ymax_{i}$, $\theta^1_{i}$ and $\theta^2_{i}$ for all subplots $i=1,\dots,1536$. $\theta^1$ and $\theta^2$ had no effect on yield, but they were spatially correlated with $\beta^{i}. This scenario was meant to reflect the lack of scientific consensus about which variables explain heterogeneous yield response to N, making it likely that the researcher would include irrelevant variables like $\theta^1$ and $\theta^2$ in real-world data analysis.

@@

CNN leaves room for researchers to determine its architecture. We used a slightly modified version of one of the multi-stream CNN architectures proposed by @barbosa2020modeling, called "Late Fusion." Briefly stated, explanatory variables except N rate (e.g. $\alpha^{i,j}$, $\beta^{i,j}$, etc.) entered the model as a set of $6 \times 6$ element rasters.The input size of the N rate is treated as $1 \times 1$ because it is spatially homogeneous within each subplot. First, each input was connected to an independent convolutional layer with eight $3 \times 3$ filters each with stride one, followed by a $2 \times 2$ max-pooling layer with stride two. Then, a fully-connected rectified linear unit (ReLU) layer with sixteen neurons was added after each max-pooling layer, followed by a single ReLU neuron. Finally, multiple neurons were concatenated and fed to the fully-connected ReLU layer with sixteen neurons, followed by an output with a linear activation function. @barbosa2020modeling demonstrated that among several CNN architectures, this architecture modeled crop yield response to N rate management best. CNN was implemented in *Python* v3.7.6 using *Pytorch* v1.7.0 [@paszke2017automatic]. The Adam optimizer [@kingma2014adam] was used with a learning rate of $0.001 \%$ (default value). To avoid over-fitting, early stopping was used to monitor validation loss with a ten epochs of patience.

Let $\mathbf{\Omega^{i}}$ denote a list of subplot-level explanatory variables (soil/field characteristics). Further, let $\hat{g}_m(N,\mathbf{\Omega})$ denote an estimated yield response function, estimated by model $m$ ($m$ = RF, BRF, or CNN). Then, yield at $i$ from model $m$ is $\hat{g}_m(N,\mathbf{\Omega^{i}})$. The site-specific EONR for each model can then be found by solving the following profit maximization problem for all $i$ for each of the models:

\begin{equation}
\hat{N}_{opt}^{i} = \operatorname*{argmax}_{N}(p \hat{g}_m(N, \mathbf{\Omega^{i}}) - w N), (\#eq:eqn8)
\end{equation}

@@

Unlike RF, BRF, and CNN, CF estimates the impact of a binary treatment. In our context, CF estimates changes in yields (i.e., $\hat{\tau}_{N^{con} \rightarrow N^{tre}}(\mathbf{\Omega^{i}})$) caused by changes in N application rates from one experimental rate (control N rate, i.e., $N^{con}$) to another experimental rate (treatment N rate, i.e., $N^{tre}$). Since we have five N application rates, four "experiments" (treatments) are identified based on four ($N^{con}$, $N^{tre}$) combinations. A possible grouping to make such pairwise N application rate combinations, which we call the "CF-base," used $(N_1, N_2),(N_1, N_3),(N_1, N_4)$, and $(N_1, N_5)$. In this approach, the treatment effect is always estimated against the lowest N rate ($N_1$) as $N^{con}$. The `grf` package (version 1.2.0 [@tibshirani2018package]) in R was used for CF estimation. Just like RF and BRF, hyperparameters were tuned using cross-validatiaon.

For estimating site-specific EONR, the trained CF-base were used to predict site-specific changes in yields resultant from changes in application rates from $N_1$ to $N_m$ $(m\in\{2,3,4,5\})$. Let $\Delta Y_{N_1 \rightarrow N_m}(\mathbf{\Omega^{i}})$ be the the estimated treatment effect of changing subplot $i$'s application rate from $N_1$ to $N_m$ for site $i$.

@@

\begin{equation}
\Delta Y_{N_1 \rightarrow N_m}(\mathbf{\Omega^{i}}) = \hat{\tau}_{N_{1} \rightarrow N_{m}}(\mathbf{\Omega^{i}}). (\#eq:eqn9)
\end{equation}

@@

The accuracy of EONR estimation was judged based on RMSE ($kg/ha$) of the EONR estimation against the true EONR at the subplot level. Using the EONR estimates, we also calculated profit-deficits ($\$/ha$) (i.e., $\hat{\pi}_{def}$) we calculated relative to the true maximum profit at the subplot level, defined as profit under the true yield response functions evaluated at $N^{opt}^{i}$. For RF, BRF, and CNN, the accuracy of yield prediction was judged based on the RMSE ($kg/ha$) of the yield prediction against the true yields at the subplot level.

>>>>

\begin{equation}
f(N_{i}: \alpha_{i,j}, \beta_{i,j}, ymax_{i,j}) = ymax_{i,j}(1-exp(\alpha_{i,j} + \beta_{i,j} \cdot N_{i})) + \varepsilon_{i,j}. (\#eq:eqn6)
\end{equation}

The first term on the right-hand side of Eq. \@ref(eq:eqn6) is the deterministic yield component. The variable $N$ represents the nitrogen fertilizer application rate (kg/ha) which the trial designs varied by plot. Also in Eq. \@ref(eq:eqn6), $\alpha$, $\beta$, and $ymax$ are spatially varying parameters that govern how yield level is determined in conjunction with the applied N rate. They reflect local cell-level production conditions, such as soil characteristics and topography. For example, $\alpha$, which represents what the yield would be when no nitrogen is applied in conjunction with $ymax$, may reflect soil organic matter and residual nitrogen from previous year's production. $ymax$, which determines the maximum yield attainable, may include other micro nutrients like phosphorus, soil depth among others. $\beta$ can be interpreted as reflecting how efficiently the crop uses N to create grain mass, and might depend on soil properties that affect the availability of applied N fertilizer to the crop, say due to N immobilization, leaching, denitrification, and volatilization [@johnson2005nitrogen; @alva2006nitrogen]. The error term of the yield $\varepsilon$ represents the composite of all the unobserved factors that affect crop yield beyond $\alpha$, $\beta$, $ymax$ and $N$. Each cell $(i,j)$'s "true" optimal N application rate, denoted $N^{opt}_{i,j}$, was calculated by solving a profit maximization problem by cell assuming that the prices of corn and N fertilizer were $\$0.138/kg$ and $\$1.323/kg$, respectively. These prices are used consistently for our simualtions. See Appendix \@ref(par-gen) for further details on how the paremeters were generated and sample maps of the parameters and true economically optimal N rates.

@@

In subsequent modeling scenarios, we made the simulated experimental data more realistic by varying the degree to which the researcher understood what the needed covariates' values were those in the data set. In the second scenario, denoted "abytt," we introduced additional variables $\theta^1$ and $\theta^2$, and the researcher's data set contained values of $\alpha_{i}$, $\beta_{i}$, $ymax_{i}$, $\theta^1_{i}$ and $\theta^2_{i}$ for all subplots $i=1,\dots,1536$. $\theta^1$ and $\theta^2$ had no effect on yield, but they were spatially correlated with $\beta_{i}$. This scenario was meant to reflect the lack of scientific consensus about which variables explain heterogeneous yield response to N, making it likely that the researcher would include irrelevant variables like $\theta^1$ and $\theta^2$ in real-world data analysis.

@@

CNN leaves room for researchers to determine its architecture. We used a slightly modified version of one of the multi-stream CNN architectures proposed by @barbosa2020modeling, called "Late Fusion." Briefly stated, explanatory variables except N rate (e.g. $\alpha_{i,j}$, $\beta_{i,j}$, etc.) entered the model as a set of $6 \times 6$ element rasters.The input size of the N rate is treated as $1 \times 1$ because it is spatially homogeneous within each subplot. First, each input was connected to an independent convolutional layer with eight $3 \times 3$ filters each with stride one, followed by a $2 \times 2$ max-pooling layer with stride two. Then, a fully-connected rectified linear unit (ReLU) layer with sixteen neurons was added after each max-pooling layer, followed by a single ReLU neuron. Finally, multiple neurons were concatenated and fed to the fully-connected ReLU layer with sixteen neurons, followed by an output with a linear activation function. @barbosa2020modeling demonstrated that among several CNN architectures, this architecture modeled crop yield response to N rate management best. CNN was implemented in *Python* v3.7.6 using *Pytorch* v1.7.0 [@paszke2017automatic]. The Adam optimizer [@kingma2014adam] was used with a learning rate of $0.001 \%$ (default value). To avoid over-fitting, early stopping was used to monitor validation loss with a ten epochs of patience.

Let $\mathbf{\Omega_{i}}$ denote a list of subplot-level explanatory variables (soil/field characteristics). Further, let $\hat{g}_m(N,\mathbf{\Omega})$ denote an estimated yield response function, estimated by model $m$ ($m$ = RF, BRF, or CNN). Then, yield at $i$ from model $m$ is $\hat{g}_m(N,\mathbf{\Omega_{i}})$. The site-specific EONR for each model can then be found by solving the following profit maximization problem for all $i$ for each of the models:

\begin{equation}
\hat{N}^{opt}_{i} = \operatorname*{argmax}_{N}(p \hat{g}_m(N, \mathbf{\Omega_{i}}) - w N), (\#eq:eqn8)
\end{equation}

@@

Unlike RF, BRF, and CNN, CF estimates the impact of a binary treatment. In our context, CF estimates changes in yields (i.e., $\hat{\tau}_{N^{con} \rightarrow N^{tre}}(\mathbf{\Omega_{i}})$) caused by changes in N application rates from one experimental rate (control N rate, i.e., $N^{con}$) to another experimental rate (treatment N rate, i.e., $N^{tre}$). Since we have five N application rates, four "experiments" (treatments) are identified based on four ($N^{con}$, $N^{tre}$) combinations. A possible grouping to make such pairwise N application rate combinations, which we call the "CF-base," used $(N_1, N_2),(N_1, N_3),(N_1, N_4)$, and $(N_1, N_5)$. In this approach, the treatment effect is always estimated against the lowest N rate ($N_1$) as $N^{con}$. The `grf` package (version 1.2.0 [@tibshirani2018package]) in R was used for CF estimation. Just like RF and BRF, hyperparameters were tuned using cross-validatiaon.

For estimating site-specific EONR, the trained CF-base were used to predict site-specific changes in yields resultant from changes in application rates from $N_1$ to $N_m$ $(m\in\{2,3,4,5\})$. Let $\Delta Y_{N_1 \rightarrow N_m}(\mathbf{\Omega_{i}})$ be the the estimated treatment effect of changing subplot $i$'s application rate from $N_1$ to $N_m$ for site $i$.

@@

\begin{equation}
\Delta Y_{N_1 \rightarrow N_m}(\mathbf{\Omega_{i}}) = \hat{\tau}_{N_{1} \rightarrow N_{m}}(\mathbf{\Omega_{i}}). (\#eq:eqn9)
\end{equation}

@@

The accuracy of EONR estimation was judged based on RMSE ($kg/ha$) of the EONR estimation against the true EONR at the subplot level. Using the EONR estimates, we also calculated profit-deficits ($\$/ha$) (i.e., $\hat{\pi}_{def}$) we calculated relative to the true maximum profit at the subplot level, defined as profit under the true yield response functions evaluated at $N^{opt}_{i}$. For RF, BRF, and CNN, the accuracy of yield prediction was judged based on the RMSE ($kg/ha$) of the yield prediction against the true yields at the subplot level.
